---
title: "STA 104 - Homework 1"
author: "Filip Wilhelm Sjostrand"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# QUESTION 3 
## a) Find the 95% confidence interval for the median
**j:** $nq-z\sqrt[]{nq(1-q)}$

**k:** $nq+z\sqrt[]{nq(1-q)}$  

### Where*:

n: The sample size

q: The quantile of interest. For a median, we will use q = 0.5.

z: The z-critical value

We round j and k up to the next integer. The resulting confidence interval is between the j:th and k:th observations in the ordered sample data.

#### *Source: 
Zach. (2021). How to Find a Confidence Interval for a Median (Step-by-Step), STATOLOGY, Available Online: [https://www.statology.org/confidence-interval-for-median/](https://www.statology.org/confidence-interval-for-median/) [Accessed 24 September 2022].


```{r}
# Step 1: Find the Sample Median
rainFall <- c(21.3, 28.8, 17.6 ,23.0, 27.2, 28.5, 32.8, 28.2, 25.9, 22.5, 27.2,
              33.1, 28.7, 24.8, 24.3, 27.1, 30.6, 26.8, 18.9, 36.3, 28.0, 17.9,
              25.0, 27.5, 27.1, 32.1, 28.0, 30.9, 20.0, 20.2, 33.5, 26.4, 30.9,
              33.2)

RainSort <- sort(rainFall)
n <- length(rainFall)
MidVal <- n/2
SampMed <- (RainSort[MidVal]+RainSort[MidVal])/2 # Sample Median


# Step 2: Find j and k
q <- 0.5
z <- 1.96 # For 95% confidence

j <- floor(n*q - z*sqrt(n*q*(1-q))) # Rounds down

k <- ceiling(n*q + z*sqrt(n*q*(1-q))) # Rounds up


# Step 3: Find the Confidence Interval
LowBound <- RainSort[23]
UppBound <- RainSort[32]
```

### Answer:

The confidence interval is **[`r LowBound`, `r UppBound`]**

## b) Make 9O% confidence intervals for the 20th and 80th percentiles


